\clearpage
\newpage
\onecolumn

\setcounter{section}{0}
 \setcounter{subsection}{0}
 \def\thesection{Appendix \Alph{section}}
 \def\thesubsection{Appendix \Alph{section}.\arabic{subsection}}

\begin{landscape}

\section{Semi-Structured Interview and Expert Interview Guidelines}
\subsection{Semi-Structured Interview Guideline (in German)}

\input{interviews/guideline_interview}


\end{landscape}
\newpage

\twocolumn

\subsection{Expert Interview Guideline}

\input{interviews/guideline_expert_interview}


\section{Transcribed Interviews}

\subsection{Transcription System}
The interviews and the focus group were transcribed the following these rules (Rules from Kuckarz \cite{kuckartz2007} with modifications):
\begin{enumerate}
    \item The transcription is literal. Dialects are not transcribed.
    \item Punctuation and language are modified to match grammar and syntax of the german language.
    \item All personal details and mentions are removed and anonymized to prevent re-identification.
    \item Pauses and breaks are marked with ellipses (\dots).
    \item Agreeing sounds like ``Mhms'', ``Ahas'', etc. of the interviewer are not transcribed if they did not interrupt the interviewee.
    \item Interjections of the other person are in brackets.
    \item Supporting or clarifying sounds of the interviewee like laughing or sighing are noted in brackets.
    \item Passages of the interviewing person are denoted with ``I:'', passages of the interviewed person with a distinct abbreviation like ``P1:''.
\end{enumerate}

\subsection{Demo and Introduction to the Application}

In each interview, the interviewer introduced and demoed the application to the interviewed person. This exemplary introduction was transcribed from the interview with participant 1. In order to retain brevity, it is omitted in the other transcriptions.

\input{interviews/introduction_interviews}

\subsection{Participant 1}

\input{interviews/interview_p1}

\subsection{Participant 2}

\input{interviews/interview_p2}

\subsection{Participant 3}

\input{interviews/interview_p3}

\subsection{Participant 4}

\input{interviews/interview_p4}

\subsection{Participant 5}

\input{interviews/interview_p5}

\subsection{Participant 6}

\input{interviews/interview_p6}

\section{Transcribed Expert Interviews}

\subsection{Interview Tobias Heide}

\input{interviews/interview_e1}

\subsection{Interview Carsten Ke{\ss}ler}

\input{interviews/interview_e2}