\section{Evaluation}
\label{chap:evaluation}
Subsequently to \hyperref[chap:methodology]{section \ref{chap:methodology}}, which covered applied evaluation techniques, this section will present findings inferred from the replies recorded in the semi-structured interviews (\hyperref[sub:ev_interviews]{sub-section \ref{sub:ev_interviews}}), expert interviews (\hyperref[sub:ev_expert_interviews]{sub-section \ref{sub:ev_expert_interviews}}) and focus group (\hyperref[sub:ev_focus]{sub-section \ref{sub:ev_focus}}).

\subsection{Semi-structured Interviews}
\label{sub:ev_interviews}
Two of the eight participating respondents were female. Mean age of participants was 40.38 with a standard deviation of 17.82. Except two, all respondents are involved in the organization of the sustainability project. The other two are members of a citizen foundation in M{\"u}nster. As the interviews were anonymized, participants and will be denoted as P1 to P8 in the following sections.\\
As mentioned in \hyperref[chap:methodology]{section \ref{chap:methodology}}, the responses were categorized to emphasize and expose important information. The categorized are ``Benefits and goals of public deliberation'', ``Current citizen-initiative communication means'', ``Attributes of spatial discussion platforms, which support dialogs and public deliberation'' and ``Other use cases for spatial discussion platforms''.\\
Transcriptions of all interview responses can be found in \ref{appendix:interviews}.

\textit{Benefits and goals of public deliberation}\\
According to the respondents, public deliberation is a step to self-determined organization of society (P2, P3, P5) and a method to inform oneself as a citizen (P1). One of the most important aspects is the ability of citizens to actively shape their environment through participation (P2, P3, P5, P6, P7). Public deliberation allows to pass on experience (P4, P8) and fosters discussions (P1, P2). P5 and P7 indicated that deliberation strengthens the public thus allowing to become an opposition to powerful corporations and the government. Generally, public deliberation is understood as an act of doing useful things together.

\textit{Current citizen-initiative communication means}\\
Currently, internal communication in initiatives is done mostly through mailing lists, e-mails and meetings. P2, P5, P7 and P8 name the use of Internet pages with simple lists and texts to transport information to the public. A wiki web application was mentioned by P7. Newspaper advertisements as medium were named by P2 and P6. In addition to that, brochures and flyers were used (P4, P8). According to P6 online PDF documents are also used. Telephone meetings and Trello\footnote{\url{https://trello.com/}} or other project management software were named by P5 and P3 respectively. P2 added the use of paper maps as means to transport information. Knowledge of online tools for creating interactive maps was generally low. P2 and P7 showed knowledge about the approach of the ``Nexthamburg'' project.

\textit{Attributes of spatial discussion platforms, which support dialogs and public deliberation}\\
Nearly all respondents mentioned the ability to inform the public through spatial discussion platforms as useful functionality of \textit{DialogMap}. The ability to gain a spatial overview of the topic (P1, P3, P4, P5, P7) and a sense of proximity to projects (P7, P8) were also mentioned as advantageous attributes for public deliberation. Visualization of proximity was an additional incentive for users to participate (P3, P4, P5, P7). According to the respondents, it is more likely that someone participates in his or her neighborhood than somewhere else.\\
From the point of view of most of the respondents (P2, P3, P5, P6, P7), possible discouraging factors for participation through spatial discussion platforms are low usability, application errors and unstructured discussions. Additionally, low application speed as impediment for participation through discussion platforms were named by P6. P7 also mentioned no initial content as impeding factor. P1 thinks citizens are too lazy to discuss public matters in online discussion forums.\\
On the other hand, communication, as well as acquaintances, ideas and execution of ideas are believed to be sparked through the creation of spatial references in discussions (P2, P5). The same assertion were made about the collection of data about public activities, problems and projects. Digitization of projects through the \textit{DialogMap} prototype is said to inspire more contributions and discussions (P2, P8). The ubiquitousness of Internet use is believed to allow the general public to engage others in discussions (P2, P5). P5 emphasized the asynchronous character of Internet based discussions as positive. It allows citizens to engage each other detached from time and place.\\
Then again, concerns about excluding groups without Internet access were raised by P1, P2 and P7. According to P7, even without the use of Internet technologies, public deliberation is highly socially selective. In addition to that, P2, P3, and P4 raised concerns about privacy and unclear data protection. P4 and P7 believed the requirement to create user accounts in the \textit{DialogMap} prototype could dampen spontaneous contributions. P4 named especially citizens above the age of 50. In his opinion these groups are more privacy aware, thus refusing registration in order to participate. If the context, in which the application is deployed and the reuse of contributions are not clearly communicated, acceptance will be generally low (P2, P8). Low acceptance induced by low publicity was also named by P2, P3, P5, and P7. Generally, the graphic representation of objects in the \textit{DialogMap} are considered useful (P2, P7, P8).\\
Displaying markers and polygons on a map allows users to easily grasp the density of projects in different areas (P7). P7 also mentioned that unpleasing visual appearance could deter users from contributing. Additionally, interactivity enables users to understand the connection between a project and its location (P4, P6).\\
Citizens with an initial low interest in topics of citizen initiatives are discouraged through complicated user interfaces (P5). Possible monetary costs of deployment and maintenance as negative factors for online spatial discussion platforms were mentioned by P4. P3 thinks, citizen initiatives have low experience in the use of Internet applications to support public deliberation. P2 stated citizen initiatives had no access and expertise to use Internet applications and especially spatial applications for their deliberation activities.

\textit{Other use cases for spatial discussion platforms}\\
The use of a spatial discussion platform as an organizational tool is mentioned by most of the respondents. Conveyance of information of all sort was named by P2, P5 and P6. P7 came up with the idea to display open data produced by governments through a spatial discussion platform. P2 also added spatial discussion platforms as suitable for self-reflection in political initiatives.

\subsection{Expert Interviews}
\label{sub:ev_expert_interviews}
As mentioned earlier, expert interviews were conducted with Carsten Ke{\ss}ler and Tobias Heide. Statements will be denoted as E1 and E2 respectively. Transcriptions can be found in \ref{appendix:expert-interviews}.\\
Regarding knowledge of existing spatial discussion platforms, E2 mentioned ArguMap.\footnote{short for argumentation map}. E1 counted emergency response maps to the class of spatial discussion platforms. He stated that bijectivity between features and objects supports mutual understanding of what discussions are about. He later stated that he observed the same bijectivity in the \textit{DialogMap} prototype.\\
``Literally everything''\footnote{``E1: Ich k{\"o}nnte mir das relativ grob strukturiert eigentlich {\"u}berall vorstellen.''} could be the use case for linking discussions with spatial features according to E1. Public participation and involving citizens in planning processes are named by E2 for use cases.\\
The idea to include spatial references in discussion contributions in public deliberation was received positively by both experts. Applications that use this concept could be used in conjunction with petition portals (E1), town hall meetings and public postings (E2). E2 remarked that the use of online applications could backfire so that non-technical persons or groups could be excluded. In his experience, usability should be prioritized in the design of spatial discussion platforms to avoid a steep learning curve.\\
The presentation of the developed prototype was generally well received by both experts. The design decision to omit spatial features created as replies, was described as good for clarity (E1) and as a good idea to prevent visual clutter (E2). E1 added that through this, the focus is on the creator of the first contribution.\\
According to E1, the two-way highlighting implemented in the prototype is an important aspect for comprehensibility of relationships between spatial features and textual contributions. E2 adds that it helps users to see and understand the relationships.\\
The ability to filter and sort contributions was deemed as less important with only few contributions by E1. The amount options users can choose from, could be confusing for users. E2 indicated that the functions were implemented as he were expecting them. He added, a spatial filter functionality could complement the filtering functions.\\
The compose and reply forms were conceived as generally good and working as expected. E2 denoted that the categories are not explained well enough and found multiple categories confusing. In the same sense, the creation of spatial references was found unintuitive by both experts. The icons for enable the referencing actions are not self-explanatory. E2 added that suggesting locations by geocoding would be helpful for users.\\
Although the feature for favoring contributions is a good measure of the popularity of a contribution, no incentives for using it exist (E1, E2). Additionally, the button is not accessible immediately and not self-explanatory (E1). E2 thought the ability to favor contributions is a good idea.\\
Registration and authentication works as expected for both experts. The social login was called ``state-of the-art'' by E1. In this sense, E2 believed third-party authentication mechanisms lower participation barriers for users.\\
In the opinion of both experts, spatial dialogs will be aided through the use of the \textit{DialogMap} prototype. E1 pointed out that a discussion tool is only as good as its moderation tools. In his opinion, a real dynamic discussion is generally difficult through text-based technology (also E2). With the small sidebar, the \textit{DialogMap} prototype leads the focus more to the map, not to the discussion. E2 added that more evaluation is needed to answer this question thoroughly.\\
To create incentives to use the favor function, E1 suggested to implement user accessible favorite-lists. He adds that there is no functionality for users to only display unread contributions or find all contributions of him or herself in the \textit{DialogMap} prototype. Furthermore, E2 missed the ability for searching spatially.\\

\subsection{Focus Group}
\label{sub:ev_focus}
Two of the three participants of the focus group discussion were female. Age of participants ranged from 53 to 60 with a standard deviation of 3.61 (mean 56). They were recruited through meetings in context to the sustainability project in Münster. Statements of the participants will be labeled as F1 to F3. Transcription of the focus group session can be found in \ref{appendix:focus-interviews}.\\
All participants indicated, that, after a short phase of becoming familiar the user interface of the prototype worked as expected. F3 noted an inconsistency when trying to edit an own contribution (To edit, users have to click ``Reply'' first). The workflow for creating spatial references and hyperlinks worked as participants intended and was generally well received.\\
The participants noted they really liked the concept of a decentralized information source where everyone with an account can add their information. Contrary to this, F1 thought the fact that contributions cannot be made anonymously, could discourage the creation of contributions and replies.\\
Correspondingly, all participants expressed disbelief for general acceptance for the concept of spatially enhanced dialogs. They stated to understand the concept, but could not think of use cases where they needed spatial discussions. In the opinion of F1, online based discussion tools require too much effort in contrast to in-person meetings. F2 and F3 think information presented to the public should have a well structure, which could not be ensured over the course of a discussion. F3 could imagine tactical planning for military operations as a better use case for a geospatial discussion platform. F1 suggests to integrate more flexible background maps with information related to the context in which the platform is deployed. Other suggestions for missing features were a news feed for missed contributions since the last visit (F1), the extension of user profiles with arbitrary address data (F1) and a calendar which visualizes the contributions chronologically (F3).
